\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}

### 2022-05-10 lin-vs-cen.tex
Got rid of stupid letterpaper option

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics{figs/bad-example-1} &
      \includegraphics{figs/bad-example-2} &
      \includegraphics{figs/bad-example-3} 
    \end{tabular}
  \end{center}
  \caption{Some examples of $3$- and $4$-element vertex subsets that cannot all be contained in a single path (that must also contain other vertices not pictured).}
\end{figure}

@@

  Define a bipartite graph $H$ with parts $L(H):=X:=\varphi(V(G))\setminus \overline{X}$ and $R(H):=Y:=Q_1$.  Let $\pi:=\mu_1,\ldots,\mu_{2|K|}$ be a random permutation of $Y$.  We include an edge $\alpha\mu_i$ in $H$ if and only if
  \[
    \VE(G_r(\mu_i))\setminus\left(\bigcup_{j=1}^{i-1}\VE(G_{2r}(\mu_j))\right)
  \]
  contains some object $\nu$ of colour $\phi(\nu)=\alpha$.  Note the asymmetry here: In the first round, each element $\mu_j$ that takes part in $Q_1$ prevents elements in $\VE(G_r(\mu))$ from also being part of $Q_1$.  In the second round, $\mu_j$ defines a larger box $\VE(G_{2r}(\mu_j))$ that must be avoided when choosing a second element from $\VE(G_r(\mu_i))$, for each $i>j$.

@@

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/25)\le 2\vol{r}) \notag \\
    & \le (2t)^{2\vol{r}}(1/25)^{r^2}(1-1/25)^{t-2\vol{r}} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & \le \exp({2\vol{r}}(1+\log(2t))-t/25) \notag \\
    & \le \exp({2\vol{r}}(1+\log(2|\Phi_\alpha|))-|\Phi_\alpha|/25\vol{10r}) \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

@@

  Therefore, by \cref{weighted_lovasz}, there exists a permutation $\alpha_1,\ldots,\alpha_{2|K|}$ such that $\deg_H(\alpha)\ge {2\vol{r}}$ for each $\alpha\in X$.  On the other hand, $\deg_{H}(y)\le |\VE(G_r(y))|-1 < {\vol{r}}$ for each $y\in Y$.  Therefore, for any $A\subseteq X$, $|N_H(A)|\ge {2\vol{r}}|A|/{\vol{r}} = 2|A|$.  Therefore, by \cref{d_hall}, there is a subgraph $M$ of $H$ in which $\deg_M(\alpha)=2$ for each $\alpha\in X$ and $\deg_M(\mu)\le 1$ for each $\mu\in Y$. Each edge $\alpha\mu$ of $M$ corresponds to some $\nu\in X_\alpha\cap\VE(G_r(\mu))$, which we place in $Q_2$.

>>>>

\begin{figure}
  \begin{center}
    \begin{tabular}{ccc}
      \includegraphics{figs/bad-example-1} &
      \includegraphics{figs/bad-example-2} &
      \includegraphics{figs/bad-example-3}
    \end{tabular}
  \end{center}
  \caption{Some examples of $3$- and $4$-element vertex subsets that cannot all be contained in a single path (that must also contain other vertices not pictured).}
  \label{bad_examples}
\end{figure}

@@

  Define a bipartite graph $H$ with parts $L(H):=X:=\varphi(V(G))\setminus \overline{X}$ and $R(H):=Y:=Q_1$.  Let $\pi:=\mu_1,\ldots,\mu_{|Y|}$ be a random permutation of $Y$.  We include an edge $\alpha\mu_i$ in $H$ if and only if
  \[
    \VE(G_r(\mu_i))\setminus\left(\bigcup_{j=1}^{i-1}\VE(G_{2r}(\mu_j))\right)
  \]
  contains some object $\nu$ of colour $\phi(\nu)=\alpha$.  Note the asymmetry here: In the first round, each element $\mu_j$ that takes part in $Q_1$ prevents elements in $\VE(G_r(\mu))$ from also being part of $Q_1$.  In the second round, $\mu_j$ defines a larger box $\VE(G_{2r}(\mu_j))$ that must be avoided when choosing a second element from $\VE(G_r(\mu_i))$, for each $i>j$.

@@

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/25)\le 2\vol{r}) \notag \\
    & = \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{1}{25}\right)^x\left(\frac{24}{25}\right)^{t-x} \notag \\
    & < \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{24}{25}\right)^{t} \notag \\
    & \le (2t)^{2\vol{r}}\cdot\left(\frac{24}{25}\right)^{t} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & = \exp\big({2\vol{r}}\ln(2t)-t\ln(25/24)\big) \notag \\
    & \le \exp\left({2\vol{r}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|\ln(25/24)}{\vol{10r}}\right)
      & \text{(since $|\Phi_\alpha|/\vol{10r} \le t \le |\Phi_\alpha|$)} \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

@@

  Therefore, by \cref{weighted_lovasz}, there exists a permutation $\pi$ of $Y$ that produces a bipartition graph $H$ with $\deg_H(\alpha)\ge {2\vol{r}}$ for each $\alpha\in X$.  On the other hand, $\deg_{H}(y)\le |\VE(G_r(y))|-1 < {\vol{r}}$ for each $y\in Y$.  Therefore, for any $A\subseteq X$, $|N_H(A)|\ge {2\vol{r}}|A|/{\vol{r}} = 2|A|$.  Therefore, by \cref{d_hall}, there is a subgraph $M$ of $H$ in which $\deg_M(\alpha)=2$ for each $\alpha\in X$ and $\deg_M(\mu)\le 1$ for each $\mu\in Y$. Each edge $\alpha\mu$ of $M$ corresponds to some $\nu\in X_\alpha\cap\VE(G_r(\mu))$, which we place in $Q_2$.
